---
title: "Data analysis and modeling - Beverage study"
author: "Mauricio Alessandrello"
fig-cap-location: margin
format: 
  html:
    toc: TRUE
    toc-location: left
    cap-location: bottom
title-block-banner: "#00524e"
title-block-banner-color: "white"
---

## Summary

This document is a data analysis and modeling report of the Beverage study.

## Exploratory Data Analysis

In this section, I load and explore the dataset. 

```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(readr)
library(rstatix)
library(ggpubr)
library(lme4)
library(lmerTest)
library(knitr)

options(digits = 2)

data <- read_csv("Study Data.csv") %>% 
  rename(id = `Participant ID`, day = `Study Day`)

head(data)


```
***

Let's analyze the question of alcohol consumption before and while consuming the product.

```{r message=FALSE, warning=FALSE}
# 1 and 0 instead of 2 and 1 for "no" and "yes" answers respectively
daily_alc <- data %>% 
  drop_na() %>% 
  filter(`Question Name` == "daily_alc") %>% 
  mutate(Value = ifelse(Value == 2, 1, 0))

daily_alc_stats <- daily_alc %>% 
  group_by(Battery) %>% 
  get_summary_stats(Value, show = c("n", "mean", "sd", "se", "ci"))

kable(daily_alc_stats)


```

The proportion of individuals that don't consume alcohol during beverage is 6% higher than the proportion of individuals that don't consume alcohol pre beverage. 

```{r message=FALSE, warning=FALSE, fig.cap= "Probability of not consuming alcohol previous to and during the consumption of the beverage. p-values were obtained with a t-test."}

daily_alc %>% 
  group_by(id, Battery) %>% 
  summarise(Value = mean(Value)) %>% 
  ggplot(aes(Battery, Value))+
  geom_boxplot()+
  geom_pwc(method = "t_test")+
  theme_minimal()+
  ylab("Probability of NOT consuming alcohol")+
  theme(axis.title.x = element_blank())

```

Do this increase in the proportion of no alcohol consumers has something to do with the consumption of the beverage? If so, we should see an increase in the chances of not drinking alcohol from the day on which the beverage is consumed. To do that, we will use a generalized mixed effects linear model because it is best adapted when the outcome we want to predict is binary.

```{r}


model <- lme4::glmer(Value ~ day + (1|id),
                     na.action = na.omit,
                     family = binomial(link = "logit"),
                     data = daily_alc)

summary(model)
```

We can see from the random effects that there is variability in the baseline acounting for individual variability and that there is an increase  
