---
title: "Data analysis and modeling - Beverage study"
author: "Mauricio Alessandrello"
fig-cap-location: margin
format: 
  html:
    toc: TRUE
    toc-location: left
    cap-location: bottom
title-block-banner: "#00524e"
title-block-banner-color: "white"
---

# Summary

This document is a data analysis and modeling report of the Beverage study.

# Alcohol consumption during and pre-beverage consumption

## Exploratory Data Analysis 

In this section, I load and explore the dataset. 

```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(readr)
library(rstatix)
library(ggpubr)
library(lme4)
library(knitr)
library(performance)
library(modelsummary)


options(digits = 2)

data <- read_csv("Study Data.csv") %>%
  rename(id = `Participant ID`,
         day = `Study Day`,
         question = `Question Name`)

head(data)


```
***

Let's analyze the question of alcohol consumption before and while consuming the product.

```{r message=FALSE, warning=FALSE}
#Chane 1 and 0 instead of 2 and 1 for "no" and "yes" answers respectively,
# bev is the Battery, coded 1 for Beverage Daily and 0 otherwise.

daily_alc <- data %>%
  drop_na() %>%
  filter(question == "daily_alc") %>%
  mutate(
    Value = ifelse(Value == 2, 1, 0),
    bev = as_factor(ifelse(Battery == "Beverage Daily", 1, 0))
  )

daily_alc_stats <- daily_alc %>%
  group_by(Battery) %>%
  get_summary_stats(Value, show = c("n", "mean", "sd", "se", "ci"))

kable(daily_alc_stats)


```
***
The proportion of "no" ansewers during beverage is 6% higher than the proportion of "no" answers when pre beverage. The difference is statistically signifcant as showed by this t-test:
```{r message=FALSE, warning=FALSE}
t.test(Value ~ bev, data = daily_alc)

```
***

In the next plot, the proportion of "no" answers per individual is depicted.

```{r message=FALSE, warning=FALSE, fig.cap= "Proportion of 'no' answers per individual previous to and during the consumption of the beverage. p-values were obtained with a t-test."}

daily_alc %>%
  group_by(id, Battery) %>%
  mutate(Value = mean(Value)) %>%
  ggplot(aes(Battery, Value)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.2) +
  geom_pwc(method = "t_test") +
  theme_minimal() +
  ylab("Proportion of 'no' answers") +
  theme(axis.title.x = element_blank())



```
***
```{r warning=FALSE, message=FALSE}
# Let's do some more calculations:

a <- daily_alc %>%
  group_by(id, Battery) %>%
  summarise(Value = mean(Value))

b <- pivot_wider(a, names_from = Battery, values_from = Value)

head(b)

# corroborating t-test in chart:
t.test(b$`Beverage Daily`, b$`Pre Beverage Daily`)

# Let's calculate the diference in the proportion of no answers for each participant:
c <- b$`Beverage Daily` - b$`Pre Beverage Daily`

# Mean increase in the proportion of "no" answers during beverage.
mean(c, na.rm = TRUE)


# This confirms the values provided by the function get_summary_stats

```

## Modeling

Do this increase in the proportion of no alcohol consumers has something to do with the consumption of the beverage? If so, we should see an increase in the chances of not drinking alcohol from the day on which the beverage is consumed. To do that, we will use a generalized mixed effects linear model because it is best adapted when the outcome we want to predict is binary.


```{r message=FALSE, warning=FALSE}


model1 <- lme4::glmer(factor(Value) ~ day + (1 | id),
  na.action = na.omit,
  family = binomial(link = "logit"),
  data = daily_alc
)

summary(model1)



```
***

We can see from the random effects that there is variability in the baseline and that there is a statistically significant increase of 5% per day in the chances of not consuming alcohol. However, this model considers every day as equal, but the days when individuals are on beverage may have more weight, if the beverage is truly working. So, let's try with an interaction model. 

```{r, message=FALSE, warning=FALSE}
model2 <- lme4::glmer(factor(Value) ~ day + day * bev + (1 | id),
  na.action = na.omit,
  family = binomial(link = "logit"),
  data = daily_alc
)

summary(model2)
```
***
This model catches an effect produced by the beverage condition that is statistically significant. The interaction term has a lower level of significance.

When we compare both models, the model 2 performs better in most metrics than model 1.

```{r message=FALSE, warning=FALSE}

compare_performance(model1, model2)

```

## Prediction of Probability of not consuming alcohol during the study

In this section, we will predict the probability of not consuming alcohol during the time of the study (0 to 21 days).

```{r message=FALSE, warning=FALSE, fig.cap = "Predicted probability of not consuming alcohol during the study." }

#set with 0 to 21 days

dataset <- with(daily_alc, tibble(day = c(0:21), bev = as_factor(c(
  0, 0, 0, 0, 0, 0, 0,
  1, 1, 1, 1, 1, 1,
  1, 1, 1, 1, 1, 1, 1,
  1, 1
))))


predicted <- predict(model2, newdata = dataset, type = "response", re.form = NA)

prediction <- tibble("Predicted values" = predicted, days = c(0:21))

prediction %>%
  ggplot(aes(days, `Predicted values`)) +
  geom_point() +
  xlab("Days") +
  ylab("Predicted Probabilty of not consuming alcohol") +
  theme_minimal() +
  annotate(geom = "text", x = 15, y = 0.91, label = "Beverage consumption starts here") +
  geom_curve(aes(x = 12, y = 0.911, xend = 7, yend = 0.93),
    arrow = arrow(length = unit(0.25, "cm"), ends = "last", type = "open")
  )

```

## Conclusion

* The prpoportion of 'no' answers during beverage was 6% higher than in the pre beverage. This difference was statistically significant with p-value = 0.021.
* For those during beverage, there is an increase in the odds ratio in favor of not alcohol drinking of 2.4 (p-value = 0.024). This increase is independent of day.



# WHO 

## Exploratory Data Analysis

Brief EDA of the dataset concerning the "WHO" questions. First, let's identify the duplicate questions:

```{r message=FALSE, warning=FALSE}
#data_dup is dataframe containing all columns except the answers filtering by 
#"who" questions. If there are duplicate questions with different answers,
#we should see duplicated rows in this dataframe.

data_dup <- data %>% 
  filter(str_detect(question, "who")) %>% 
  select(id, day, Battery, question) %>% 
  drop_na()

anyDuplicated(data_dup)

```

We see that there are no duplicated rows, indicating that there are no replicate questions. Let's continue exploring the dataset.

```{r, message=FALSE, warning=FALSE}

data_who <- data %>% 
  filter(str_detect(question, "who")) %>% 
  drop_na()

head(data_who)

#Calculating the who-score for every id to obtain a single row for week.

data_who_calc <- data_who %>% 
  group_by(id, day) %>% 
  mutate(Value = sum(Value)*4) %>% 
  select(-question, -Text) 

data_who_calc <- data_who_calc[!duplicated(data_who_calc),]



data_who_stats <-  data_who_calc %>% 
  group_by(day) %>% 
get_summary_stats(Value)

kable(data_who_stats)

```
***
Evaluating the means and the confidence interavals, we can say that there is no statistically singnificant increase in the who-score. Let's evaluate this with a t-test:

```{r, message=FALSE, warning=FALSE}
# Convert the dataframe to a wide format to obtain one row per id:

data_who_wide <- data_who_calc %>% 
  select(-Battery) %>% 
  pivot_wider(names_from = day, values_from = Value) %>% 
  drop_na()

t.test(data_who_wide$`0`, data_who_wide$`21`)

```
***
Statistically, there is no significant difference between the who-score at the beginning of the study and the final one. However WHO indicates that a 10% difference is indicative of significant difference. Let's calculate that difference:

```{r, message=FALSE, warning=FALSE}

who_score_dif <- tibble(id = data_who_wide$id,
       "1" = data_who_wide$`7` - data_who_wide$`0`,
       "2" = data_who_wide$`14` - data_who_wide$`7`,
       "3" = data_who_wide$`21` - data_who_wide$`14`,
       "total" = data_who_wide$`21` - data_who_wide$`0`)

who_score_dif_weekly <- who_score_dif %>%
  select(-total) %>% 
  pivot_longer(cols = 2:4, names_to = "week", values_to = "Value") %>% 
  mutate(week = as.numeric(week))

who_score_dif_total <- who_score_dif %>% 
  select(id, total)

```
