---
title: "Data analysis and modeling - Beverage study"
author: "Mauricio Alessandrello"
fig-cap-location: margin
format: 
  html:
    toc: TRUE
    toc-location: left
    cap-location: bottom
title-block-banner: "#00524e"
title-block-banner-color: "white"
---

## Summary

This document is a data analysis and modeling report of the Beverage study.

## Exploratory Data Analysis

In this section I load and explore the dataset. 

```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(readr)
library(rstatix)
library(ggpubr)
library(lme4)
library(lmerTest)
library(knitr)
options(digits = 2)

data <- read_csv("Study Data.csv") %>% 
  rename(id = `Participant ID`)

head(data)


```
***

Let's analyze the question of alcohol consumption before and while consuming the product.

```{r message=FALSE, warning=FALSE}

daily_alc <- data %>% 
  drop_na() %>% 
  filter(`Question Name` == "daily_alc") %>% 
  mutate(Value = ifelse(Value == 2, 1, 0))

daily_alc_stats <- daily_alc %>% 
  group_by(Battery) %>% 
  get_summary_stats(Value, show = c("n", "mean", "sd", "se", "ci"))

kable(daily_alc_stats)


```

We can see that there is an increase of 6% in the proportion of individuals that do not consume alcohol while consuming the beverage. Confidence intervals of both groups don't overlap, indicating statistical significance. This can also be seen graphically:

```{r message=FALSE, warning=FALSE, fig.cap= "Probability of not consuming alcohol previous to and during the consumption of the beverage. p-values were obtained with a t-test."}

daily_alc %>% 
  group_by(id, Battery) %>% 
  summarise(Value = mean(Value)) %>% 
  ggplot(aes(Battery, Value))+
  geom_boxplot()+
  geom_pwc(method = "t_test")+
  theme_minimal()+
  ylab("Probability of NOT consuming alcohol")+
  theme(axis.title.x = element_blank())

```

Let's examine the initial proportion 

```{r}
mean(daily_alc$Value == 0)
```

